Variational Monte Carlo on bosonic systems

### 2018-03-25 project_1/article/article.tex

        Scanning the entire space of variational parameters can be a somewhat
        despairing task. To aid us in the search we therefore implement the
        method of gradient descent\footnote{Also known as steepest descent,
        easily confused with the steepest descent method for integation.}. In
        any optimum we generally have that the derivative should equal zero. In
        our case, we want to find the minimum of the expected local energy with
        respect to some variational parameter $\alpha$, so that
        \begin{equation}
            \nabla_\alpha \expval{E_L(\mathbf{R}; \alpha)} = 0.
        \end{equation}
        The foundation for  the method of gradient descent lies in the fact that
        some analytic \footnote{We use the term analytic quite recklessly.}
        function $F(\alpha)$ decreases fastest in the direction of
        $-\nabla_\alpha F(\alpha)$. This means that we could eventually get to
        an ensuing minimum by trailing along the pathway of the following
        recursive relation,
        \begin{equation}
            \alpha_{k+1} = \alpha_k - \gamma \nabla_{\alpha} F(\alpha_k).
        \end{equation}
        If $\gamma$ is small enough we should have $f(\alpha_k) \geq
        F(\alpha_{k+1})$ for all $k$ and hopefully we will experience
        convergence to the desired minimum.

        For our special case : INSERT DERIVATION OF LOCAL ENERGY derivative wrt alpha??

@@

    \begin{figure}
        \centering
            \includegraphics[width=250px]{../data/figures/problem_c1.pdf}
            \caption{As the number of Monte Carlo cycles increases the standard deviation of the local energies
            decrease. This effect is stronger when the importance sampling scheme is employed.}
            \label{fig:problem_c1}
    \end{figure}

    \begin{figure}
        \centering
            \includegraphics[width=250px]{../data/figures/problem_c2.pdf}
            \caption{The absolute difference between the expected energy and the exact energy for both brute force
            and importance sampling. }
            \label{fig:problem_c2}
    \end{figure}

@@

    \autoref{tab:imp_time_step} shows the standard deviations and acceptance
    ratios for different time steps. The simulations in this table are from the
    same span $\alpha \in [0.3, 07]$ as in \autoref{tab:1D1N_analytic} and
    \autoref{tab:1D1N_numeric}. The optimum energies for $\alpha = \frac{1}{2}$
    gives the same result for importance sampling as for the naïve way of
    sampling. Notice that by lowering the time step, the uncertainty starts to
    decrease and then starts to increase again. Moreover, the acceptance ratios
    increases as the time step decreases.

    \subsubsection{Test of Gradient descent}
        Before moving on to the interacting, more interesting systems, it would
        be beneficial to test the gradient descent implementation now that we
        are certain that it moves towards the optimum variational parameter
        $\alpha = \frac{1}{2}$.

    \subsection{Interacting elliptical harmonic oscillator}
        We now look at the interacting, elliptical, harmonic oscillator with the
        scaled system shown in \autoref{eq:scaled_hamiltonian} and scaled single
        particle functions from \autoref{eq:scaled_spf} with the Jastrow factor.
        Due to similarities with the spherical harmonic oscillator system we
        make a guess that the true minimum of this new system should be situated
        close the minimum of the previous system. Choosing seven values for
        $\alpha \in [0.2, 0.7]$ we run importance sampling on this system for $N
        = \{10, 50, 100\}$ particles in $d = 3$ dimensions using $2^{21}$ Monte
        Carlo cycles\footnote{This turned out to be a little bit overkill due to
        the fast convergence of importance sampling.} and an additional $10\%$
        of the Monte Carlo cycles for thermalization of the system prior to
        doing any sampling. We have used $\delta t = 0.1$ with $a_{\text{ho}} =
        0.0043$ and $\lambda = \beta = \sqrt{8}$.

        We have sampled the expectation value for the local energy, the standard
        deviation using the blocking method, the acceptance ratio of the method
        and the CPU time spent for each value $\alpha$. The results for $N = 10$
        particles are shown in \autoref{tab:10_interacting}, $N = 50$ in
        \autoref{tab:50_interacting} and finally for $N = 100$ in
        \autoref{tab:100_interacting}.

@@

>>>>

        Scanning the entire space of variational parameters can be a somewhat despairing task. To aid us in the search we
        therefore implement the method of gradient descent\footnote{Also known as steepest descent, easily confused
        with the steepest descent method for integation.}. In any optimum we generally have that the derivative should equal
        zero. In our case, we want to find the minimum of the expected local energy with respect to some variational
        parameter $\alpha$, so that
        \begin{equation}
            \nabla_\alpha \expval{E_L(\mathbf{R}; \alpha)} = 0.
        \end{equation}
        The foundation for  the method of gradient descent lies in the fact that some analytic
        \footnote{We use the term analytic quite recklessly.} function $F(\alpha)$ decreases fastest in the direction
        of $-\nabla_\alpha F(\alpha)$. This means that we could eventually get to an ensuing minimum by trailing along the
        pathway of the following recursive relation,
        \begin{equation}
            \alpha_{k+1} = \alpha_k - \gamma \nabla_{\alpha k} F(\alpha_k).
        \end{equation}
        If $\gamma$ is small enough we should have $f(\alpha_k) \geq F(\alpha_{k+1})$ for all $k$ and hopefully we will
        experience convergence to the desired minimum.

        For our special case we only have one variational parameter, $\alpha$.  The expression for the derivative is,
        \begin{equation}
            \frac{d E_L}{d\alpha} =
                2\left( \expval{\frac{1}{\psi_T}\frac{d\psi_T}{d\alpha} E_L} - \expval{\frac{1}{\psi_T} \frac{d\psi_T}{d\alpha}}\expval{E_L} \right).
        \end{equation}

        INSERT DERIVATION OF LOCAL ENERGY??

@@

            \begin{figure}
                \centering
                    \includegraphics[width=250px]
                    {../data/figures/problem_c1.pdf}
                    \caption{As the number of Monte Carlo cycles increases the
                    standard deviation of the local energies decrease. This
                    effect is stronger when the importance sampling scheme is
                    employed.}
                    \label{fig:problem_c1}
            \end{figure}

            \begin{figure}
                \centering
                    \includegraphics[width=250px]
                    {../data/figures/problem_c2.pdf}
                    \caption{The absolute difference between the expected energy
                    and the exact energy for both brute force and importance
                    sampling. }
                    \label{fig:problem_c2}
            \end{figure}

@@

            \autoref{tab:imp_time_step} shows the standard deviations and
            acceptance ratios for different time steps. The simulations in this
            tabel are from the same span $\alpha \in [0.3, 07]$ as in
            \autoref{tab:1D1N_analytic} and \autoref{tab:1D1N_numeric}. The
            optimum energies for $\alpha = \frac{1}{2}$ gives the same result
            for importance sampling as for the naïve way of sampling. Notice
            that by lowering the time step, the uncertainty starts to decrease
            and then starts to increase again. Moreover, the acceptance ratios
            increases as the time step decreases.

        \subsubsection{Test of Gradient descent}

            Before moving on to the interacting, more interesting systems, it
            would be beneficial to test the gradient descent implementation now
            that we are certain that it moves towards the optimum variational
            parameter $\alpha = \frac{1}{2}$.

    \subsection{Interacting elliptical harmonic oscillator}
        We now make a change to the simple systems in the previous section in
        order to make everything a bit more interesting. The potential in the
        one-body Hamiltonian is perturbed in the $z$-direction according to
        \autoref{eq:scaled_hamiltonian} by setting $\lambda = \sqrt{8} \approx
        2.82843$. This corresponds to enclosing particles of the simulation in
        an elliptical trap instead of a spherical trap.

        We also choose a new trial wavefunction with the elliptical gaussian
        single particle functions in \autoref{eq:scaled_spf} and the Jastrow
        factor. The hard sphere radius is set to $a/a_{\text{ho}} = 0.0043$.

        Due to similarities with the spherical harmonic oscillator
        system we make a guess that the true minimum of this new system should
        be situated close to the minimum of the previous system. Choosing seven
        values for $\alpha \in [0.2, 0.7]$ we run importance sampling on this
        system for $N = \{10, 50, 100\}$ particles in $d = 3$ dimensions using
        $2^{21}$ Monte Carlo cycles\footnote{This turned out to be a little bit
        overkill due to the fast convergence of importance sampling.} and an
        additional $10\%$ of the Monte Carlo cycles for thermalization of the
        system prior to doing any sampling. We have used $\delta t = 0.1$ yields
        a high acceptance ratio, but slower convergence.

        In \autoref{tab:10_interacting} the system is set to have $N = 10$
        interacting particles. Notice first that the energy is higher in
        the interacting system than for a comparable non-interacting system
        (\autoref{tab:initial_optimum_comparison}).  Moreover, the standard
        deviation gets lower as the system is close to an optimal variational
        parameter.

@@

        The result of simulations with a higher number of particles, $N=50$ and
        $N=100$, are  shown in \autoref{tab:50_interacting} and
        \autoref{tab:100_interacting} respectively. We observe that the energy
        for systems of interacting particles increases rapidly as the size of
        the system increases.
